---
title: "OpenCPU guide"
author: "Peter"
date: "2/6/2018"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(eval=FALSE, echo=FALSE)
```

## What this doc is about

This document outlines how to work with OpenCPU on a local machine
when developing application workflows (R) and user interface
(HTML, JS; we won't use CSS for now).
We will assume that all R functionality that is needed for the app
already exists, i.e. we will not cover R package development
at this stage. We will also use local single-user server and 
the public ocpu cloud server,
i.e. not cover remote hosting and server administration.
We will purely focus on the interaction
between the the UI front-end and the OpenCPU server back-end.

## Prerequisites

* R installed,
* **opencpu** package installed via `install.packages("opencpu")`,
* text editor (RStudio's editor is fine, or Atom, Brackets, etc.),
* a modern browser with developer tools (e.g. Chrome) with internet connection.

## Hello World!

Create a file called `hello-world.html` with the following content:

```
<!DOCTYPE html>
  <head>
    <script src="http://code.jquery.com/jquery-1.11.1.min.js"></script>
    <script src="http://cdn.opencpu.org/opencpu-0.4.js"></script>
  </head>
  <body>
    <h1 id="output"></h1>
    <button id="submitbutton" type="button">Click me!</button>
    <script>
      ocpu.seturl("http://cloud.opencpu.org/ocpu/library/base/R")
      var mydata = "Hello World!";
      $("#submitbutton").on("click", function(){
        ocpu.rpc("print", {
            x : mydata
          }, function(output){
            $("#output").text(output);
          });
      });
    </script>
  </body>
</html>
```

Now let's inspect the code line-by-line: the following skeleton
is standard HTML, the head declares everything that we can't see (meta info),
the body has everything that we see in the browser. This is a pretty bare-bone
file for the sake of simplicity:

```
<!DOCTYPE html>
  <head>
  
    ...

  </head>
  <body>

    ...

  </body>
</html>
```

The 2 lines in the head will load prerequisites for the
OpenCPU JavaScript API to work, namely jQuery (which is 
a JS library that defines many of the behaviour os the
HTML elements, e.g. events like clicking a button):

```
    <script src="http://code.jquery.com/jquery-1.11.1.min.js"></script>
    <script src="http://cdn.opencpu.org/opencpu-0.4.js"></script>
```

Within the body we have 2 elements, `h1` (level 1 heading)
with nothing in between its opening `<h1>` and closing `</h1>` tags,
and with ID `"output"` (usually there is some text between the tags).

The other element is a button, opening and closing tags,
with ID `"submitbutton"`:

```
    <h1 id="output"></h1>
    <button id="submitbutton" type="button">Click me!</button>
```

The last piece in our file is a script inside the html body
between the tags `<script> ... </script>`.
The point of placing this script at the very end of the file is
that jQuery and opencpu JS files load first,
and the rest of the page is also loaded.
This way the JS actions defined in the script (so called handlers)
will be bound to the html elemnts (referenced by their IDs).

This line sets the OpenCPU URL to the cloud server.
This is called cross-domain OpenCPU request, or [CORS](https://www.opencpu.org/jslib.html#lib-cors).
So this is where OpenCPU will look for the R code to use.
We set it to **base** R for now (`/library/base/R`, no backslash at the end).
The base of the URL (`http://cloud.opencpu.org/ocpu`) is the root
of the OpenCPU server.

```
      ocpu.seturl("http://cloud.opencpu.org/ocpu/library/base/R")
```

Next, we define a variable called `mydata`:

```
      var mydata = "Hello World!";
```

In the remaining part of the script, we use the `ocpu.rpc` function.
Its first argument is the name of the R function we want to call,
this must be exported by the R package in question (**base** in this case).
We'll use the `print` function.

The second argument is a list of arguments passed to the R function,
in JSON format.

The third argument is a so called callback function
that defines what to do with the R function's return value.
In thus case, it takes the output and displays it as text within the `h1` 
element.

Note also that all this `ocpu.rpc` is wrapped
within `$("#submitbutton").on("click", function(){ ... });`
that binds the click on the `button` element to the server side call:

```
      $("#submitbutton").on("click", function(){
        ocpu.rpc("print", {
            x : mydata
          }, function(output){
            $("#output").text(output);
          });
      });
```

Next we open the html file in Chrome, then open developer tools
from its menu (View > Developer > Developer Tools).
Clicking on the Console tab, we should be seeing something like this:

```
Setting path to CORS server http://cloud.opencpu.org/ocpu/library/base/R
opencpu-0.4.min.js:27 Path updated. Available objects/functions:
^
~
<
<<-

...

xzfile
zapsmall
```

By clicking the button, we should now see the 'Hello World!' header.

## Using single user server

Load the **opencpu** package and start 

```{r}
library(opencpu)
ocpu_start_server()
```

Now open the page `http://localhost:5656/ocpu`
in the browser. You should see a test page where you can make
requests to the local OpenCPU server. If that is the case,
let's modify the html page and replace the line
`ocpu.seturl("http://cloud.opencpu.org/ocpu/library/base/R")`
with
`ocpu.seturl("http://localhost:5656/ocpu/library/base/R")`.
Then reload the page, click the button and see what happens.
It should be the same as the cloud server.
Hit Ctrl+C or Esc to stop the local server.

## Exercise 1

Change `mydata` to a numeric vector
(`var mydata = [1,2,3,4,5,6,7,8,9,10,11,12,13,14,15];`)
and have the R function `sd` in the **stats** package to calculate
its standard deviation. Put that inside of a paragraph
instead of a heading (use the `<p> ... </p>` tag pair) following the button.

## Exercise 2

Use the `rnorm` function from the **stats** package
to generate 10 random numbers with mean -1 and standard deviation 3,
similarly to `rnorm(n = 10, mean = -1, sd = 3)`. Again, use
paragraph instead of heading to display the results.
